Numerical Stability of Detonations in White Dwarf Simulations

### 2018-07-17 papers/ignition/paper.tex
Clear the results sections

Another example of a failure mode comes from the timestep limiting described in
\ref{sec:timestep}. The timestep will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion $\Delta t \leq f_{X_k} \tau_{X_k}$
when the estimates for $\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions over the timestep. However,
this can assumption will sometimes fail during periods of runaway burning when the
rate of change of these quantities is highly nonlinear. We may not want to neglect
the errors caused by this approximation because they may build up over an extended
period of nonlinear evolution and perhaps substantially change the final results.

@@

We implement the white dwarf collision problem in \castro\footnote{The main results
in this paper were generated using the 17.12 releases of \castro, \amrex, and
\microphysics.} in the \wdmerger\ problem
directory. We prepare the domain in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along a coordinate axis.
In this paper we focus only on head-on collisions of two 0.64 $\msolar$ white dwarfs
composed of equal parts carbon and oxygen by mass, at an initially uniform temperature
of $T = 10^7$ K. This choice is for comparison to earlier studies; the code does have
the ability to vary the impact parameter of the collision and the mass and composition
of the WDs.

We have added to the \wdmerger\ problem setup the ability to take advantage of 
the 2D cylindrical ($R-z$) coordinate system evolution in \castro, and we use it
in this study, because the axisymmetry inherent to the cylindrical coordinate system
lends itself well to the head-on collision problem. In this coordinate system, we
align the WDs along the $z$-axis (which is analogous to the $x$-axis in Cartesian
evolution), with the center of the WDs at $R = 0$. The un-simulated $\phi$ dimension
then would extend the WDs through a $2\pi$ revolution. The domain has width $4 \times 10^{9}$
cm along the $z$ axis, and is half as wide along the $R$ axis. The resolution is equal
in both dimensions, so that there are twice as many zones in the $z$ dimension in
a uniform grid.

We terminate the simulation at $t = 9$ seconds, as we have found that for this
particular setup, all of the significant burning has completed by this time\footnote{In
some simulations we were mainly interested in seeing when the detonation happened, and
we terminated the simulation at that point. We show nucleosynthesis results only for
those simulations that had completed nuclear burning.}. At
this point, the total energy on the domain has become positive, typically about
$10^{51}$ ergs, and has stopped increasing. This means that the system has become
unbound due to nuclear energy release, and that no further meaningful nuclear energy
generation is occurring. In our discussions below, it helps to have some other
anchoring points in time when significant events occur. We arbitrarily define the
collision time, $t_{\rm c}$, as the time when the density at the center of the grid
first reaches $10^2\ \text{g\,/\,cm}^3$. This is reached only when the WDs meet at the
center of the domain. We define the detonation time, $t_{\rm d}$, as the time when the
temperature first reaches $4.5 \times 10^9\ \text{K}$. This is also an arbitrary
threshold, but we have found empirically that this temperature is only reached in
the collision when a detonation is formed.

The region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g\,/\,cm}^{3}$ whose composition and temperature is the
same as that of the WDs. Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g\,/\,cm}^{3}$ and $T < 10^8\ \text{K}$. We have confirmed that
the results do not depend on this choice. A sponge is applied to all gas of density
lower than $1\ \text{g\,/\,cm}^{3}$. The sponge damps out the velocity in these zones
(see Paper I) to prevent spurious numerical error in low-density gas from dragging
down the global timestep. The density floor for the simulation is set
at $10^{-5}\ \text{g\,/\,cm}^{3}$ and the temperature floor is $10^7$ K.

We collect information at the end of every timestep about the total amount of
all of the species on the grid (in solar masses), and we use the mass of $^{56}$Ni
generated as a metric for this test, as this is the parameter most directly related
to the observable quantities of interest for Type Ia supernovae. Since, as we will see
below, not all of our simulations detonate at a density high enough to generate substantial
amounts of nickel, we employ the mass-weighted average atomic mass number on the domain,
$\bar{A}$, as a more general metric describing the outcome of the simulation. A
final relevant metric is the total amount of energy $E$ released by the detonation.

Plotfiles of relevant quantities are created every 0.01 seconds.
The timesteps are limited to ensure that every multiple of this plotfile interval is
reached exactly, which means an effective maximum timestep of 0.01 seconds. Timesteps
are by default limited only by the hydrodynamic stability constraint and this plotfile
interval constraint, except where noted otherwise.

For a collision of two 0.64 $\msolar$ WDs with 128 zones in the radial dimension,
corresponding to a uniform spatial resolution of 312.5 km, the collision time is
$t_{\rm c} = 4.16\ \text{s}$, and the detonation time is $t_{\rm d} = 7.43\ \text{s}$.
The peak density at the detonation time is $2.3 \times 10^7\ \text{g\,/\,cm}^3$.
Approximately $0.21\ \msolar$ of nickel and $0.46\ \msolar$ of silicon is created,
corresponding to an $\bar{A}$ of about 29. Approximately $1.3 \times 10^{51}$
ergs of energy are released. Simulations with resolutions in this few hundred kilometer
range have been a common choice in earlier papers on white dwarf mergers and collisions
due to computational constraints; our goal is to see how higher resolution runs differ.

%==========================================================================
% Resolution Dependence
%==========================================================================
\section{Spatial Resolution}
\label{sec:spatialresolution}

Now we turn to an examination of the effect of spatial resolution on the detonation.
The first question we want to answer is: at reasonable uniform spatial resolution, is
the simulation converged when a detonation occurs? If the answer is no, then the
follow-up question is: can adaptive mesh refinement help achieve a converged simulation?

To answer these questions we performed a series of simulations at varying base
resolutions, augmented with additional runs with AMR. In each case we ran the
simulation to a point just after the burning started, but before a detonation
had occurred. Then we stopped
the run and created a checkpoint. Starting from that checkpoint, we did a simulation
at varying levels of refinement based on the unstable burning criterion. We will refer
to the refinement through the refinement factor, $r$. This factor is the jump in
resolution between the coarse grid and the finest grid. The uniform base grid case is $r = 1$.
Meanwhile, $r = 16$ means a uniform coarse grid plus two levels of 4x refinement.

At high resolution the collision time, $t_{\rm c} = 4.6\ \text{s}$, is later than the
collision time at low resolution, due to a sharper view of the contact point. However,
for the same reason, the detonation time is significantly earlier at $t_{\rm d} = 6.62\ \text{s}$,
as the peak density reaches the key density range of $10^7\ \text{g\,/\,cm}^3$ significantly earlier,
and this is what triggers the burning sufficient to generate an earlier detonation.
Approximately speaking, the later $t_{\rm d}$ is, the more
nickel is produced. As time passes, more white dwarf material slams into the center, compressing
and heating up the gas. Both effects are favorable to production of nickel. So, a later $t_{\rm d}$
implies that more material can be converted to nickel.

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.425]{{{plots/amr}}}
  \caption{Nickel production as a function of the simulation resolution. The solid
           black curve connects the points at uniform resolution. For each uniform
           resolution simulation, we include additional simulations using the adaptive
           mesh refinement criterion described in \autoref{sec:unstable_burning}.
           The horizontal axis is the effective number of zones per dimension on the
           finest level, which is the number of zones per dimension that would exist
           if the finest resolution was uniformly applied on the grid.
           \label{fig:amr}}
\end{figure}

@@

This research has made use of NASA's Astrophysics Data System 
Bibliographic Services. In addition, this research has made use
of the AstroBetter blog and wiki.

\software{\amrex\ (\url{https://github.com/AMReX-Codes/Amrex}),
          \castro\ \citep{castro} (\url{https://github.com/AMReX-Astro/Castro}),
          \wdmerger\ \citep{wdmergerI} (\url{https://github.com/AMReX-Astro/wdmerger}),
          GCC (\url{https://gcc.gnu.org/}),
          python (\url{https://www.python.org/}),
          matplotlib \citep{matplotlib} (\url{http://matplotlib.org/}),
          yt \citep{yt} (\url{http://yt-project.org/})}

\facilities{OLCF, NERSC}

>>>>

Another example of a failure mode comes from the timestep limiting described in
\autoref{sec:timestep}. The timestep will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion $\Delta t \leq f_{X_k} \tau_{X_k}$
when the estimates for $\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions over the timestep. However,
this can assumption will sometimes fail during periods of runaway burning when the
rate of change of these quantities is highly nonlinear. We may not want to neglect
the errors caused by this approximation because they may build up over an extended
period of nonlinear evolution and perhaps substantially change the final results.

@@

Plotfiles of relevant quantities are created every 0.01 seconds.
The timesteps are limited to ensure that every multiple of this plotfile interval is
reached exactly, which means an effective maximum timestep of 0.01 seconds. Timesteps
are by default limited only by the hydrodynamic stability constraint and this plotfile
interval constraint. For simulations with resolution better than 64 km, we use a CFL
number of 0.5. For the simulations with lower resolution, we use a CFL number of 0.01.
This is because at low resolution, the timestep is so large that the detonation can
form in a single timestep and represent such a large change that the burning integration
fails (\autoref{sec:retries}). The timestep retry mechanism can handle this gracefully,
but the subcycled timestep necessary to do so is usually so small that the full level
timestep ends up being so long that it just makes more sense to avoid this situation
entirely and use smaller level advance timesteps.

@@

The authors thank Chris Malone and Don Willcox for useful discussions
on the nature of explosive burning, and Doron Kushnir for providing
clarification on the nature of the burning limiter used in \cite{kushnir:2013}.

This research has made use of NASA's Astrophysics Data System 
Bibliographic Services.
